---
title: 'Homework 1'
subtitle: "PSTAT 122: Design and Analysis of Experiments" 

author: "Spring 2026"
format: 
 pdf:
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
    geometry: margin=0.7in 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span>  
</div>
- Aldo Nunez (A399512)


:::

:::{.callout-tip}
# Instructions

* Please write your **full name** and **perm number** legibly on the front page of your homework sheet.

* This homework must be completed and submitted **individually**. Collaboration is allowed for discussion, but each student must submit their own work.

* Solve all homework problems. You may use `R` for your calculations and analysis, but you must include all `R` code used.  

* If you prefer to solve some questions by hand, you may do so. Please scan or take a clear photo of your handwritten work and embed it as an image under the corresponding question. Ensure that your handwriting is neat and legible.  

* Your reasoning must be clear and complete. To receive full credit, provide sufficient explanations alongside any code and results.

* Ensure that all `R` code, mathematical formulas, and workings are presented clearly and appropriately.

* All figures should be numbered, and axes must be labeled. 

* Please use the provided `HW 1 S26.qmd` file to type your solutions and submit the completed assignment as a PDF file. You can utilize `RStudio` for this purpose. For guidance, refer to the [Tutorial: Hello, Quarto](https://quarto.org/docs/get-started/hello/rstudio.html)).

* Submit your solution via **Gradescope** as **a single PDF file**.


:::

:::{.callout-caution}
# Due Date
**Due Date:** April 10, 2026, 11:59 PM
:::

\newpage

:::{.callout-important}
## Question 1  

Suppose that you want to investigate the factors that potentially affect students’ performance in an online course.

a) What would you use as a response variable in this experiment? How would you measure the response?

b) List all of the potential sources of variability that could impact the response.

c) Complete the first three steps of the guidelines for designing experiments in Section 1.4 (slide 44).
::: 



`r bfcolor("ANSWERS TO QUESTION 1:", "red")` \

a) The response variable in this experiment would be student performance. You would measure he response based on what grade they receive in the course.

b) Some potential sources of variability are how long a student spent studying, how motivated they were during the course, and how good their internet was

c) 1. Randomly assign students to different versions of the course
   2. Apply each treatment to multiple student instead of one
   3. Group students based on year and major
  
  
\newpage
:::{.callout-important}
## Question 2 
   Let $X \sim N(0,1)$ and $Y = X^2$
   
   - (a) Find Cov$(X,Y)$.  
   - (b) Find $P(Y > 1 \mid X = 1.5)$ and $P(Y > 1 \mid X = -0.3)$.  
   - (c) What dou you conclude from (a) and (b)?  
     
   **Note**: *We have shown in class that if two random variables are independent then they are uncorrelated. Here we see that the converse is not true.*
   
:::  
   
 
`r bfcolor("ANSWERS TO QUESTION 2:", "red")` \
(a)
$$
Cov(X,Y) = Cov(X, X^2) = \mathbb{E}[X * X^2] - \mathbb{E}[X]\mathbb{E}[X^2] = \mathbb{E}[X^3] - \mathbb{E}[X]\mathbb{E}[X^2]
$$
Since $X \sim N(0,1)$: $E[X] = 0$, $E[X^2] = 1$, and $E[X^3] = 0$

Therefore, $$Cov(X,X^2) = 0 -(0)(1) = 0$$

(b)
$$\mathbb{P}(Y >1 | X=1.5) = \mathbb{P}(2.25 >1) =1$$
$$\mathbb{P}(Y >1 | X=-0.3) = \mathbb{P}(0.09 >1) =0$$


\newpage
:::{.callout-important}
## Question 3  


The following data represent the daily number of steps (in thousands) taken by 25 individuals participating in a fitness study.

$$
6.5\quad 7.8\quad 8.2\quad 5.9\quad 9.1\quad 10.3\quad 7.2\quad 8.7\quad 6.8\quad 9.5
$$
$$
7.9\quad 8.4\quad 6.1\quad 10.0\quad 9.8\quad 7.5\quad 8.9\quad 6.7\quad 7.3\quad 9.2
$$
$$
8.0\quad 7.6\quad 6.9\quad 9.7\quad 8.5
$$

a) Find the sample mean, median, interquartile range, and sample standard deviation.  

b) What percentage of values lie within 2 standard deviations of the mean?  

c) Find the Z-score for the value 5.9.  

d) Construct a boxplot and comment on the distribution. 

:::

`r bfcolor("ANSWERS TO QUESTION 3:", "red")` \

(a)
```{r}
steps <- c(6.5, 7.8, 8.2, 5.9, 9.1, 10.3, 7.2, 8.7, 6.8, 9.5, 7.9, 8.4, 6.1, 10.0, 9.8, 7.5, 8.9, 6.7, 7.3, 9.2, 8.0, 7.6, 6.9, 9.7, 8.5)

print(paste("sample mean:", mean(steps)))
print(paste("median:", median(steps)))
print(paste("IQR:", IQR(steps)))
print(paste("sample standard deviation:", sd(steps)))
```
(b)
```{r}
lower_range <- mean(steps) - 2*sd(steps)
upper_range <- mean(steps) + 2*sd(steps)

ctr <- 0
for(i in 1:length(steps)){
  if(steps[i] >= lower_range && steps[i] <= upper_range){
    ctr <- ctr + 1
  }
}

percentage <- (ctr / length(steps)) * 100
cat("Percentge is:", percentage)
```

(c)
```{r}
z <- (5.9 - mean(steps)) / sd(steps)
cat("Z-score for 5.9:", z)
```
(d)
```{r}
boxplot(steps, main="Boxplot of steps", ylab="Values", col = "lightblue")
```
The data seems to be very compact 
\newpage
:::{.callout-important}
## Question 4 

A dataset consists of $n$ observations with mean $A$ and standard deviation $B$. A new dataset is created using the transformation:

$$Y = aX + b$$

a) Express the mean and standard deviation of $Y$ in terms of $A$, $B$, $a$, and $b$. Clearly justify your answer.

b) Suppose a dataset has sample mean $\bar{x} = 15$ and standard deviation $s_x = 4$ with $n = 12$. After applying the transformation, the new dataset has mean $\bar{y} = 5$ and standard deviation $s_y = 8$.

- Find all possible values of $a$ and $b$.  
- Interpret what happens when $a$ is negative.

Show all your calculations.
:::

`r bfcolor("ANSWERS TO QUESTION 4:", "red")` \


a) Mean: $E[Y] = E[aX+b] = aE[X] + b = aA +b$
  Standard Deviation: $SD(Y) = \sqrt{Var(aX+b)} = \sqrt{a^2Var(X)} = |a|B$
  
b) $15a +b = 5$ and $|a|4 = 8$
a = 2 or -2

if a = 2: 15(2) + b = 5 $\implies$ b = -25
if a = -2: 15(-2) + b = 5 $\implies$ b = 35


A negative a reverses the order of the data but still scales the spread by the absolute value

\newpage
:::{.callout-important}
## Question 5  

Is there sufficient evidence to indicate that the average number of hours per week spent studying is greater for students majoring in engineering than for those majoring in business?

**Data Summary:**

| Major | Engineering | Business |
|----------|------------------|-------------|
| Sample size (n) | 100 | 90 |
| Sample mean (x̄) | 18.5 hours | 15.2 hours |
| Sample standard deviation (s) | 5.4 hours | 4.8 hours |

a) State the null and alternative hypotheses.

b) What is the appropriate rejection region for an $\alpha = 0.05$ level test?
	
c) Calculate the observed value of the appropriate test statistic.

d) What is your conclusion?

e) What is a practical reason for the conclusion you reached in part (d)?

::: 
`r bfcolor("ANSWERS TO QUESTION 5:", "red")` \


a) $$H_0 : \mu_E = \mu_B \\
    H_A: \mu_E > \mu_B$$


b) Reject $H_0$ if $Z > 1.645$

c) $$Z = \frac{18.5-15.2}{\sqrt{\frac{5.4^2}{100} + \frac{4.8^2}{90}}} = \frac{3.3}{0.74} \approx 4.46$$

d) Since 4.46 > 1.645, we reject the null hypothesis since there is highly significant evidence that Engineering majors study more

e) A reason for the conclusion that Engineering majors study more is that Engineering is often viewed as more difficult than business
\newpage

:::{.callout-important}
## Question 6 


The state of California is working hard to ensure that all elementary-age students whose native language is not English become proficient in English by the sixth grade. Their progress is monitored each year using the California English Language Development Test. The results for two school districts in southern California for the 2003 school year are given in the following table.



| District     | Riverside | Palm Springs |
|--------------|-----------|--------------|
| Students     | 6124      | 5512         |
| % Fluent     | 40%       | 37%          |

Do the data indicate a significant difference in the 2003 proportions of students who are fluent in English for the two districts? Use $\alpha = 0.01$.
:::

`r bfcolor("ANSWERS TO QUESTION 6:", "red")` \

```{r}
res <- prop.test(x = c(0.40*6124, 0.37*5512),
                 n = c(6124, 5512),
                 conf.level = 0.99)
res
```
$Z = \frac{0.40 - 0.37}{\sqrt{0.3858(1 - 0.3858) \left( \frac{1}{6124} + \frac{1}{5512} \right)}} = \frac{0.03}{0.009} \approx 3.33$

For $alpha = 0.01$, the critical value is $Z = 2.578$. Since 3.33 > 2.576, we reject the null hypothesis. There is a significant difference in fluency proportions between the two districts