---
title: "pstat126_homework3"
author: "Melody Su"
date: "2026-05-28"
output:
  pdf_document: default
  html_document: default
---

```{r}
install.packages("car", dependencies = TRUE)
```

```{r}
data <- read.csv("Dataset for Homework 3-1.csv")
data$Exam_Level <- as.factor(data$Exam_Level)
data$Exam_Anxiety <- as.factor(data$Exam_Anxiety)

full_model <- lm(
  Exam_Score ~ Hours_Studied +
    Phone_Usage +
    Assignments_Completed +
    Exam_Level +
    Exam_Anxiety,
  data = data
)

summary(full_model)
```

## 1. Variable Selection

Use the AIC/BIC method to select the most appropriate predictors for the model

```{r}
# AIC
aic_model <- step(full_model, direction = "backward")
summary(aic_model)
AIC(aic_model)
```

```{r}
# BIC
n <- nrow(data)
bic_model <- step(
  full_model,
  direction = "backward",
  k = log(n)
)
summary(bic_model)
```

The best model according to both AIC and BIC was:

$$\text{Exam Score ~ Hours Studied + Assignments Completed + Exam Level}$$

The variables **Phone Usage** and **Exam Anxiety** were removed because they did not improve the model sufficiently.

```{r}
final_model <- lm(
  Exam_Score ~ Hours_Studied +
    Assignments_Completed +
    Exam_Level,
  data = data
)
AIC(final_model)
BIC(final_model)
```

\
Best model statistics:

- AIC = 753.3955

- BIC = 769.0266

## 2. Multicollinearity Check

Apply the Variance Inflation Factor (VIF) technique to detect and remove multicollinearity among the predictors.

```{r}
library(car)
vif(aic_model)
```

The VIF values for **Hours Studied**, **Assignments Completed**, and **Exam Level** are all below 5, indicating that multicollinearity is not a concern. Therefore, no variables need to be removed based on the VIF analysis.

## 3. Final Model Specification

Write down the final reduced model obtained after applying these techniques.

$$\widehat{\text{Exam Score}} = \text{-7.4503 + 13.1683(Easy) + 10.4826(Medium) + 4.2816(Hours Studied) + 7.3101(Assignments Completed)}$$

## 4. Interpretation of Estimates

Provide a statistical interpretation of the coefficients output (estimates, standard error, t-value) in the final model.

#### [**Hours Studied**:]{.underline} 

For every additional hour studied, the expected exam score increases by about 4.28 points, holding all other variables constant

The standard error is small (0.361), meaning the estimate is precise

The large t-value (11.845) and very small p-va;ue indicate that **Hours Studied** is highly significant.

#### [Assignments Completed:]{.underline}

For every additional assignment completed, the expect exam score increases by about 7.31 points, holding other variables constant.

The large t-value (10.169) shows strong statistical significance.

#### [Exam Level:]{.underline}

Compared to a difficult exam:

- students taking an easy exam score about 13.17 points higher on average

- students taking a medium exam score about 10.48 points higher on average

Both coefficients are statistically significant

#### [Intercept:]{.underline}

The intercept represents the predicted exam score when all predictors equal zero and the exam is difficult.

Although the intercept is statistically significant, it may not have a strong practical interpretation because some predictor values may not realistically equal zero.

## 5. Hypothesis Testing

### (a) Partial Significance Tests

Explain the significance of individual coefficients based on their p-values. (This test assesses the significance of individual predictors in the model by checking whether each coefficient is significantly different from zero).

For each predictor:

$$H_0: \beta_j = 0\\
H_A : \beta_j \neq 0$$

Hours Studied: significant

Assignments Completed: significant

Exam Level (Easy): significant

Exam Level (Medium): significant

Since all p-values are below 0.05, all predictors in the final model are statistically significant.

### (b) Global Significance Test

Interpret the F-statistic to assess the overall significance of the model, (This test evaluates whether at least one predictor in the model is significant).

$$F = 69.57$$

with p-value:

$$2.14 \times 10^{-27}$$

Hypothesis:

$$H_0: \beta_1 = \beta_2 = \cdots = \beta_p = 0\\
H_A : \text{At least one } \beta_j \neq 0$$

Since the p-value is extremely small, we reject the null hypothesis.

Conclusion: The regression model is globally significant, meaning at least one predictor is useful for explaining exam scores.

## 6. Model Performance

### (a) interpret $R^2$ and Adjusted $R^2$ to assess model fit

Model statistics:

- $R^2 = 0.746$

- Adjusted $R^2 = 0.735$

Interpretation:

- The model explains approximately 74.6% of the variability in exam scores.

- The adjusted $R^2$ is slightly smaller because it accounts for the number of predictors used.

- Since both values are relatively high, the model provides a good fit to the data.

### (b) Residual Standard Error (RSE)

The Residual Standard Error is:

$RSE \approx 7.37$

Interpretation:

- The RSE measures the typical prediction error of the model.

- On average, the predicted exam scores differ from the actual scores by about 7.37 points.

- A smaller RSE indicates better prediction accuracy.

Overall, the model performs well and explains a substantial portion of the variation in exam scores.
